---
title: "Derivation: exact mean and variance of the bivariate cross nearest-neighbour G function"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Derivation: exact mean and variance of the bivariate cross nearest-neighbour G function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette proves the closed-form mean and variance of the bivariate cross
nearest-neighbour statistic \eqn{G_{ij}} under a disjoint label-permutation null,
again without reference to code.  The univariate result of the companion
vignette is recovered as the special case \eqn{i = j}.

## 1. The null model

The locations \eqn{X = \{x_1, \dots, x_N\}} are fixed.  Two marks (or two
phenotypes) \eqn{i} and \eqn{j} are present with observed counts \eqn{n_i} and
\eqn{n_j}, and no location carries both marks.  The bivariate CSR null permutes
the labels among the locations while preserving those counts, so

\deqn{S_i \sim \mathrm{Uniform}\{T \subset X : |T| = n_i\},}

and, conditionally on \eqn{S_i},

\deqn{S_j \sim \mathrm{Uniform}\{T \subset X \setminus S_i : |T| = n_j\}.}

This requires \eqn{n_i + n_j \le N}, which the package checks.

The statistic measures how close type-\eqn{i} points are to type-\eqn{j}
points,

\deqn{G_{ij}(r) \;=\; \frac{1}{n_i} \sum_{p \in S_i}
  \mathbf{1}\!\left\{ d(p, S_j) \le r \right\},
  \qquad d(p, S_j) = \min_{q \in S_j} \|x_p - x_q\|,}

with the convention that an empty minimum is \eqn{+\infty} (so a type-\eqn{i}
point with no type-\eqn{j} point within \eqn{r} contributes zero).

## 2. Indicator decomposition and the conditioning lemma

Put, for each location \eqn{p},

\deqn{A_p(r) = \{ q \neq p : \|x_p - x_q\| \le r\}, \qquad m_p(r) = |A_p(r)|,}

and

\deqn{Y_p \;=\; \mathbf{1}\{p \in S_i\} \cdot
  \mathbf{1}\!\left\{A_p(r) \cap S_j \neq \emptyset\right\}.}

Then \eqn{G_{ij}(r) = n_i^{-1} \sum_p Y_p}.  Unlike the univariate case, the
indicator involves *both* random sets.

**Lemma 4 (conditional uniformity).** Let \eqn{T \subseteq X} with
\eqn{|T| = t}, and condition on the event \eqn{T \subseteq S_i}.  Then
\eqn{S_j} is uniform over the \eqn{n_j}-subsets of \eqn{X \setminus T}.  In
particular, for any fixed \eqn{M \subseteq X \setminus T} with \eqn{|M| = m},

\deqn{\Pr\!\left(S_j \cap M = \emptyset \,\middle|\, T \subseteq S_i\right)
  \;=\; \frac{\binom{N-t-m}{n_j}}{\binom{N-t}{n_j}}
  \;=\; T_{N-t,\,n_j}(m).}

*Proof.*  The joint law of \eqn{(S_i, S_j)} is invariant under permutations of
\eqn{X} that preserve the marks, so it can be generated by ordering \eqn{X}
uniformly at random and taking the first \eqn{n_i} locations as \eqn{S_i} and
the next \eqn{n_j} as \eqn{S_j}.  Conditional on the set of locations occupying
the first \eqn{t} positions being exactly \eqn{T}, the remaining
\eqn{N - t} locations are in a uniformly random order, so positions
\eqn{t+1,\dots,t+n_j} form a uniform \eqn{n_j}-subset of \eqn{X\setminus T}.
Avoidance is then Lemma 1 applied to that ground set. \eqn{\square}

## 3. Exact mean

**Theorem 3 (bivariate mean).**

\deqn{\mathbb{E}\big[G_{ij}(r)\big] \;=\; \frac{1}{N} \sum_{p=1}^{N}
  \Big(1 - T_{N-1,\,n_j}\big(m_p(r)\big)\Big).}

*Proof.*  By linearity, \eqn{\mathbb{E}[G_{ij}(r)] = n_i^{-1} \sum_p
\Pr(Y_p = 1)}.  For fixed \eqn{p},

\deqn{\Pr(Y_p = 1) = \Pr(p \in S_i)\,
  \Pr\!\left(A_p(r) \cap S_j = \emptyset\,\middle|\, p \in S_i\right).}

Now \eqn{\Pr(p\in S_i) = n_i/N}, and Lemma 4 with \eqn{T = \{p\}} (so
\eqn{t = 1}) and \eqn{M = A_p(r) \subseteq X\setminus\{p\}} gives

\deqn{\Pr\!\left(A_p(r) \cap S_j = \emptyset\,\middle|\, p\in S_i\right)
  = T_{N-1,\,n_j}(m_p).}

Multiplying by \eqn{n_i/N} and dividing the sum by \eqn{n_i} yields
\eqn{(1/N)\sum_p (1 - T_{N-1,n_j}(m_p))}. \eqn{\square}

**Remark 5 (the mean depends only on \eqn{n_j}).** The type-\eqn{i} count
\eqn{n_i} enters Theorem 3 only through the prefactor \eqn{1/N}, so it cancels:
the expected cross-\eqn{G} curve is unchanged if \eqn{n_i} changes, as long as
\eqn{n_j}, and hence the type-\eqn{j} mark, is fixed.  This asymmetry is
intuitive: how many type-\eqn{i} points there are does not change the chance that
a given one finds a type-\eqn{j} neighbour.  It is also visible in the
implementation, where only \eqn{n_j} enters the avoidance table.

## 4. Exact second moments

**Lemma 6 (pair inclusion, bivariate).**

\deqn{\Pr(p \in S_i,\; q \in S_i) = \frac{n_i(n_i-1)}{N(N-1)} \;=:\; c_2^{(i)},}

which is Lemma 2 applied to the uniform \eqn{n_i}-subset \eqn{S_i}.

For the joint probability of two hits, define the *punctured* neighbour sets

\deqn{A_p^{\ast} = A_p(r) \setminus \{q\}, \qquad
      A_q^{\ast} = A_q(r) \setminus \{p\}.}

These are the neighbours of \eqn{p} and \eqn{q} other than each other, so both
are subsets of the ground set \eqn{X \setminus \{p,q\}} used by Lemma 4 with
\eqn{T = \{p, q\}}, \eqn{t = 2}.  Write

\deqn{u_{pq} \;=\; \big| A_p^{\ast} \cup A_q^{\ast} \big|
             \;=\; m_p^{\ast} + m_q^{\ast} - w_{pq},}

where \eqn{m_p^{\ast} = |A_p^{\ast}|}, \eqn{m_q^{\ast}=|A_q^{\ast}|} and
\eqn{w_{pq} = |A_p(r) \cap A_q(r)|}.

**Lemma 7 (two hits).**

\deqn{\mathbb{E}[Y_p Y_q] = c_2^{(i)}\,
  \Big(1 - T_{N-2,\,n_j}(m_p^{\ast}) - T_{N-2,\,n_j}(m_q^{\ast})
       + T_{N-2,\,n_j}(u_{pq})\Big).}

*Proof.*  \eqn{\{Y_p = Y_q = 1\}} requires \eqn{p, q \in S_i}, of probability
\eqn{c_2^{(i)}}, and then that \eqn{S_j} meets both \eqn{A_p(r)} and
\eqn{A_q(r)}.  Conditional on \eqn{p,q\in S_i}, Lemma 4 with
\eqn{T = \{p,q\}} says \eqn{S_j} is uniform over the \eqn{n_j}-subsets of
\eqn{X \setminus \{p,q\}}.  On that ground set the relevant neighbour sets are
exactly \eqn{A_p^{\ast}} and \eqn{A_q^{\ast}}.  By inclusion--exclusion,

\deqn{\Pr(\text{both met}) = \frac{\binom{N-2-m_p^{\ast}}{n_j}}{\binom{N-2}{n_j}}
  \!+\! \frac{\binom{N-2-m_q^{\ast}}{n_j}}{\binom{N-2}{n_j}}
  \!-\! \frac{\binom{N-2-u_{pq}}{n_j}}{\binom{N-2}{n_j}},}

which is the stated bracket. \eqn{\square}

Note that, unlike the univariate case, there is no special short-cut when
\eqn{\|x_p - x_q\| \le r}: the hit condition concerns \eqn{S_j}, not
\eqn{S_i}.  The general formula already covers that situation, because then
\eqn{q \in A_p(r)} so \eqn{m_p^{\ast} = m_p - 1}, and the corresponding
avoidance probability decreases accordingly.

**Theorem 4 (bivariate variance).**

\deqn{\mathrm{Var}\big(G_{ij}(r)\big) \;=\; \frac{1}{n_i^2}
  \left[\sum_{p=1}^{N} e_p(1 - e_p)
      + 2\sum_{p<q}\big(f_{pq} - e_p e_q\big)\right],}

with \eqn{e_p = (n_i/N)(1 - T_{N-1,n_j}(m_p))} and \eqn{f_{pq}} from Lemma 7.

*Proof.*  Identical to Theorem 2, using the decomposition
\eqn{G_{ij} = n_i^{-1}\sum_p Y_p}. \eqn{\square}

## 5. The diagonal case

When the two marks coincide, the natural reading of \eqn{G_{ii}} is the
univariate nearest-neighbour statistic computed on a uniform subset of size
\eqn{n_i}: there is no independent \eqn{S_j}.  The package implements
\eqn{G_{ii}} by delegating to the univariate estimator with \eqn{n = n_i}, and
Theorems 1--2 then apply verbatim.  This is the correct reduction because in
the univariate problem the neighbour of a marked point may be any other marked
point, whereas in Lemma 7 \eqn{S_j} and \eqn{S_i} are disjoint.

## 6. Numerical illustration

For small patterns the bivariate null can be enumerated exhaustively: for each
of the \eqn{\binom{N}{n_i}} choices of \eqn{S_i}, run over all
\eqn{\binom{N-n_i}{n_j}} choices of \eqn{S_j} from the remaining locations.  With
\eqn{N = 10}, \eqn{n_i = 4} and \eqn{n_j = 3} this is
\eqn{\binom{10}{4}\binom{6}{3} = 210 \times 20 = 4200} configurations.

```{r bivariate enumeration, message = FALSE}
suppressPackageStartupMessages(library(permfreeG))
set.seed(11)
N <- 10; ni <- 4; nj <- 3
x <- stats::runif(N); y <- stats::runif(N)
radii <- seq(0, 1.2, by = 0.15)

D <- as.matrix(stats::dist(cbind(x, y)))
ci <- utils::combn(N, ni)
Gsim <- NULL
for (j in seq_len(ncol(ci))) {
  Si <- ci[, j]
  rem <- setdiff(seq_len(N), Si)
  cj <- utils::combn(length(rem), nj)
  for (k in seq_len(ncol(cj))) {
    Sj <- rem[cj[, k]]
    Gsim <- rbind(Gsim,
                  vapply(radii, function(rr)
                    mean(rowSums(D[Si, Sj, drop = FALSE] <= rr) > 0), numeric(1)))
  }
}

d <- data.frame(x = x, y = y,
                m = factor(c(rep("A", ni), rep("B", nj),
                             rep("C", N - ni - nj))))
res <- as.data.frame(exact_gcross(d, radii = radii, mark_col = "m",
                                  marks_i = "A", marks_j = "B"))

c(max_mean_error     = max(abs(res$csr_mean - colMeans(Gsim))),
  max_variance_error = max(abs(res$csr_var -
                               apply(Gsim, 2,
                                     function(z) mean((z - mean(z))^2)))))
```

Again the errors are pure floating-point round-off.

A direct check of Remark 5 is also immediate: changing \eqn{n_i} while holding
\eqn{n_j} fixed leaves the exact mean unchanged.

```{r remark-check, message = FALSE}
pp <- spatstat.geom::ppp(x, y,
                         window = spatstat.geom::owin(range(x), range(y)),
                         marks = d$m)
a <- permfreeG:::csr_cross_moments_cpp(x, y, radii, 4L, nj)
b <- permfreeG:::csr_cross_moments_cpp(x, y, radii, 5L, nj)
max(abs(a$mean - b$mean))
```
