---
title: "Derivation: exact mean and variance of the univariate nearest-neighbour G function"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Derivation: exact mean and variance of the univariate nearest-neighbour G function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette proves, without reference to any code, the closed-form mean and
variance of the univariate nearest-neighbour \eqn{G} function under the
mark-permutation null.  The bivariate case is proved in the companion vignette
"Derivation: bivariate Gcross", and the edge corrections are derived as a
supplement in "Edge corrections".

## 1. The null model and the statistic

Let \eqn{X = \{x_1, \dots, x_N\} \subset \mathbb{R}^2} be the observed spatial
locations.  These are held **fixed**.  A mark (for example a positive cell
phenotype) is assigned to exactly \eqn{n} of them, and the null hypothesis of
complete spatial randomness (CSR) for the marks says that every assignment is
equally likely.  Equivalently, the marked set

\deqn{S \sim \mathrm{Uniform}\{\textstyle T \subset X : |T| = n\}}

is a uniformly random \eqn{n}-subset of the \eqn{N} locations.

For \eqn{r \ge 0} write
\deqn{d(p, A) = \min_{q \in A} \|x_p - x_q\|}
for a point \eqn{p} and a set \eqn{A}, and define the sample nearest-neighbour
distance of \eqn{p} within \eqn{S} as \eqn{d(p, S \setminus \{p\})}.  The
univariate \eqn{G} statistic is the empirical distribution function of those
distances over the marked set,

\deqn{G_S(r) \;=\; \frac{1}{n} \sum_{p \in S}
  \mathbf{1}\!\left\{ d(p, S \setminus \{p\}) \le r \right\}.}

Throughout we use the closed convention "\eqn{\le r}", which is exactly what
spatstat.explore::Gest() computes; the difference from a strict inequality is a
null set for continuous locations.

## 2. Indicator decomposition

The key step replaces the average over \eqn{S} by a sum of point indicators.
For each location \eqn{p \in X} put

\deqn{A_p(r) = \{ q \neq p : \|x_p - x_q\| \le r\}, \qquad m_p(r) = |A_p(r)|,}

and

\deqn{X_p \;=\; \mathbf{1}\{ p \in S \} \cdot
  \mathbf{1}\!\left\{ A_p(r) \cap S \neq \emptyset \right\}.}

Then

\deqn{G_S(r) \;=\; \frac{1}{n} \sum_{p = 1}^{N} X_p.}

Indeed, the term for \eqn{p} is one precisely when \eqn{p} belongs to the marked
set and some other selected point lies within distance \eqn{r}; dividing by
\eqn{n = |S|} turns the count into the proportion of marked points that have a
neighbour.

## 3. The avoidance table

All moments below are built from one combinatorial quantity.  For integers
\eqn{B \ge s \ge 0} and \eqn{0 \le k \le B}, define

\deqn{T_{B,s}(k) \;=\; \frac{\binom{B-k}{s}}{\binom{B}{s}},
  \qquad T_{B,s}(k) := 0 \text{ whenever } B - k < s.}

\eqn{T_{B,s}(k)} is the probability that a uniformly drawn \eqn{s}-subset of a
set with \eqn{B} elements avoids \eqn{k} prescribed elements.  The package
precomputes it once per analysis from logarithms of the gamma function, which is
why no binomial coefficients are ever formed explicitly.

**Lemma 1 (avoidance).** Let \eqn{S} be a uniform \eqn{n}-subset of a set
\eqn{X} with \eqn{|X| = N}, and let \eqn{M \subset X} with \eqn{|M| = m}.  Then

\deqn{\Pr(S \cap M = \emptyset) \;=\; T_{N,n}(m)
  \;=\; \frac{\binom{N-m}{n}}{\binom{N}{n}}.}

*Proof.*  Choosing \eqn{S} uniformly is equivalent to choosing its complement
\eqn{X \setminus S}, a uniform \eqn{(N-n)}-subset, and
\eqn{S \cap M = \emptyset} holds exactly when that complement contains all of
\eqn{M}.  The number of \eqn{(N-n)}-subsets containing a fixed \eqn{m}-set is
\eqn{\binom{N-m}{N-n} = \binom{N-m}{(N-m)-(N-n)} = \binom{N-m}{n}}; dividing by
\eqn{\binom{N}{N-n} = \binom{N}{n}} gives the claim.  Both sides are zero when
\eqn{N - m < n}. \eqn{\square}

## 4. Exact mean

**Theorem 1 (mean).** With the notation of Section 2,

\deqn{\mathbb{E}\big[G_S(r)\big] \;=\; \frac{1}{N} \sum_{p=1}^{N}
  \Big( 1 - T_{N-1,\,n-1}\big(m_p(r)\big) \Big).}

*Proof.*  By linearity and the decomposition of Section 2,

\deqn{\mathbb{E}[G_S(r)] = \frac{1}{n} \sum_{p=1}^{N} \Pr(X_p = 1).}

Fix \eqn{p}.  Since \eqn{\Pr(p \in S) = n/N}, the multiplication rule gives

\deqn{\Pr(X_p = 1) = \frac{n}{N}\,
  \Pr\!\left(A_p(r) \cap S = \emptyset \,\middle|\, p \in S\right).}

Conditional on \eqn{p \in S}, the remaining members of \eqn{S} form a uniform
\eqn{(n-1)}-subset of the \eqn{N-1} locations in \eqn{X \setminus \{p\}}, and
\eqn{A_p(r)} is a fixed subset of that ground set with \eqn{m_p} elements.
Lemma 1 therefore gives

\deqn{\Pr\!\left(A_p(r) \cap S = \emptyset \,\middle|\, p \in S\right)
  = T_{N-1,\,n-1}(m_p).}

Substituting and using \eqn{(n/N)\cdot(1/n) = 1/N} completes the proof.
\eqn{\square}

Two immediate consequences are worth recording.

* The mean is **sample-specific**: it depends on the observed locations through
  the neighbour counts \eqn{m_p(r)} alone.
* It never depends on which locations happen to carry the mark, only on
  \eqn{n}; relabelling the marks leaves \eqn{\mathbb{E}[G_S(r)]} unchanged.

## 5. Exact second moments

For the variance we need both marginal and joint probabilities.

**Lemma 2 (pair inclusion).** For distinct locations \eqn{p, q},
\deqn{\Pr(p \in S,\; q \in S) = \frac{n(n-1)}{N(N-1)} \;=:\; c_2 .}

*Proof.*  There are \eqn{\binom{N}{n}} equally likely marked sets; those
containing both \eqn{p} and \eqn{q} number \eqn{\binom{N-2}{n-2}}.  The ratio is
\eqn{n(n-1)/[N(N-1)]}. \eqn{\square}

**Lemma 3 (joint emptiness).** Let \eqn{p \neq q} and suppose
\eqn{\|x_p - x_q\| > r}, so that \eqn{q \notin A_p(r)} and
\eqn{p \notin A_q(r)}.  Write \eqn{w_{pq} = |A_p(r) \cap A_q(r)|}.  Then

\deqn{\Pr\!\left(A_p \cap S = A_q \cap S = \emptyset
  \,\middle|\, p, q \in S\right)
  = T_{N-2,\,n-2}(m_p) + T_{N-2,\,n-2}(m_q)
    - T_{N-2,\,n-2}\!\left(m_p + m_q - w_{pq}\right).}

*Proof.*  Conditional on \eqn{p, q \in S}, the other marked locations form a
uniform \eqn{(n-2)}-subset of the remaining \eqn{N-2} locations.  Both
\eqn{A_p(r)} and \eqn{A_q(r)} are fixed subsets of that ground set, of sizes
\eqn{m_p} and \eqn{m_q}, with intersection of size \eqn{w_{pq}}.  The event that
\eqn{S} meets neither set is the complement of "meets \eqn{A_p}" or "meets
\eqn{A_q}", and the union has size \eqn{m_p + m_q - w_{pq}}.  Applying Lemma 1
to each of the three sets gives the stated expression. \eqn{\square}

Combining Lemmas 2 and 3,

\deqn{\mathbb{E}[X_p X_q] =
\begin{cases}
c_2, & \|x_p - x_q\| \le r,\\[4pt]
c_2\Big(1 - T_{N-2,n-2}(m_p) - T_{N-2,n-2}(m_q)
   + T_{N-2,n-2}(m_p + m_q - w_{pq})\Big), & \|x_p - x_q\| > r.
\end{cases}}

The first case is immediate: if \eqn{q} itself lies within distance \eqn{r} of
\eqn{p}, then selecting both already guarantees that each has a neighbour, so
the conditional probability of two hits is one.

**Theorem 2 (variance).**

\deqn{\mathrm{Var}\big(G_S(r)\big) \;=\; \frac{1}{n^2}
  \left[ \sum_{p=1}^{N} e_p (1 - e_p)
       + 2 \sum_{p < q} \big( f_{pq} - e_p\, e_q \big) \right],}

where \eqn{e_p = \mathbb{E}[X_p] = (n/N)\big(1 - T_{N-1,n-1}(m_p)\big)} and
\eqn{f_{pq} = \mathbb{E}[X_p X_q]} is given above.

*Proof.*  Expand the square of the sum:

\deqn{n^2 \,\mathrm{Var}(G_S) = \sum_p \mathrm{Var}(X_p)
  + 2\sum_{p<q} \big(\mathbb{E}[X_p X_q] - e_p e_q\big),}

because the indicators are Bernoulli, so \eqn{\mathrm{Var}(X_p) = e_p(1-e_p)},
and covariance is bilinear. \eqn{\square}

The result is **exact** for every finite sample: no asymptotic approximation,
no independence assumption and no simulation are used.  The only numerical
input is the avoidance table, computed in log space.

## 6. Computational cost

Computing all \eqn{m_p(r)} costs one pass over the \eqn{\binom{N}{2}} pairs per
radius.  The double sum has one term per pair as well, but each needs the
intersection size \eqn{w_{pq} = |A_p(r) \cap A_q(r)|}.  Storing each
\eqn{A_p(r)} as a packed bit set of \eqn{\lceil N/64\rceil} machine words and
using a population count makes each intersection \eqn{O(N/64)} rather than
\eqn{O(N)}, so the total work is

\deqn{O\!\left( R\, N^2 \,\big/\, 64 + R\, N^2 \right)}

for \eqn{R} radii.  The original implementation instead called a binomial
coefficient routine inside the inner loop, which is what made the exact variance
prohibitive for large \eqn{N}.

If only a rough scale is needed, setting covariance = FALSE keeps the diagonal
term and drops \eqn{2\sum (f_{pq} - e_pe_q)}; this is the "independent points"
approximation.  It costs \eqn{O(RN)} after the neighbour counts are known.

## 7. Numerical illustration

The following small example checks Theorem 1 and Theorem 2 against *exhaustive
enumeration* of all \eqn{\binom{10}{4} = 210} marked sets.

```{r enumeration, message = FALSE}
suppressPackageStartupMessages(library(permfreeG))
set.seed(42)
N <- 10; n <- 4
x <- stats::runif(N); y <- stats::runif(N)
radii <- seq(0, 1.2, by = 0.15)

# exhaustive enumeration over all C(N, n) subsets
D <- as.matrix(stats::dist(cbind(x, y)))
ci <- utils::combn(N, n)
Gsim <- matrix(NA_real_, ncol(ci), length(radii))
for (j in seq_len(ncol(ci))) {
  S <- ci[, j]
  DS <- D[S, S]; diag(DS) <- Inf
  Gsim[j, ] <- vapply(radii, function(k) mean(rowSums(DS <= k) > 0), numeric(1))
}

res <- as.data.frame(exact_gest(data.frame(x = x, y = y,
                                           m = factor(c(rep("A", n),
                                                        rep("B", N - n)))),
                                radii = radii, mark_col = "m", marks_i = "A"))

max_mean_err <- max(abs(res$csr_mean - colMeans(Gsim)))
max_var_err  <- max(abs(res$csr_var -
                        apply(Gsim, 2, function(z) mean((z - mean(z))^2))))
c(max_mean_error = max_mean_err, max_variance_error = max_var_err)
```

Both errors are at the level of floating-point round-off, confirming that
Theorems 1 and 2 reproduce the permutation distribution exactly.

```{r plot, fig.height = 4, fig.alt = "Exact CSR mean and standard deviation band overlaid on the exhaustive-enumeration mean."}
plot(radii, colMeans(Gsim), type = "l", lwd = 3, col = "grey80",
     xlab = "r", ylab = "G(r)", main = "Exact null vs exhaustive enumeration")
lines(res$r, res$csr_mean, col = "red", lty = 2)
sd <- sqrt(res$csr_var)
lines(res$r, res$csr_mean - sd, col = "red", lty = 3)
lines(res$r, res$csr_mean + sd, col = "red", lty = 3)
legend("bottomright", c("enumeration mean", "exact mean", "+/- 1 sd"),
       col = c("grey80", "red", "red"), lty = c(1, 2, 3), bty = "n")
```
