---
title: "Supplement: edge corrections for the permutation-free G function"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Supplement: edge corrections for the permutation-free G function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4)
```

This vignette supplements the two derivation vignettes.  It defines each edge
correction supported by the package, states exactly which part of its null
distribution is available in closed form, and says plainly where an
approximation enters.  The notation follows "Derivation: univariate Gest".

## 1. Why a correction is needed

A point \eqn{p} near the boundary of the observation window has no neighbours
outside it.  Its observed nearest-neighbour distance is therefore biased upward,
and the raw empirical \eqn{G} curve underestimates the true one.  All
corrections below trade a small amount of information for reduced bias; they are
different estimators, not different null models.  Crucially, under the
mark-permutation null *the locations do not move*, so every correction's null
distribution is still induced purely by the random choice of \eqn{S}.

Let

\deqn{b_p = d(p, W^c)}

denote the distance from location \eqn{p} to the boundary of the window
\eqn{W}, and let \eqn{u_p(r) = \mathbf{1}\{b_p > r\}} mark the points that are
still "at risk" at radius \eqn{r}.

## 2. Correction none (raw)

The raw estimator is exactly the statistic of Theorem 1,

\deqn{\hat{G}_{\mathrm{none}}(r) = G_S(r)
  = \frac{1}{n}\sum_{p\in S} X_p,}

so its null mean and variance are given exactly by Theorems 1 and 2.  This is
the only fully exact combination in the package: mean **and** variance, for both
\eqn{G} and \eqn{G_{ij}}, with no approximation of any kind.

## 3. Correction rs (reduced sample / border)

Only points farther than \eqn{r} from the boundary contribute:

\deqn{\hat{G}_{\mathrm{rs}}(r) = \frac{Z(r)}{Y(r)}, \qquad
  Y(r) = \sum_{p=1}^{N} u_p(r)\,\mathbf{1}\{p \in S\}, \qquad
  Z(r) = \sum_{p=1}^{N} u_p(r)\, X_p .}

Write \eqn{\bar{u}(r) = N^{-1}\sum_p u_p(r)} and
\eqn{a_p = 1 - T_{N-1,n-1}(m_p)}.

**Mean of the denominator.**  \eqn{Y} is a simple random sample sum, so

\deqn{\mathbb{E}[Y] = n\,\bar{u}, \qquad
  \mathrm{Var}(Y) = \frac{n(N-n)}{N-1}\,\bar{u}\,(1-\bar{u}).}

**Mean of the numerator.**

\deqn{\mathbb{E}[Z] = \frac{n}{N}\sum_{p=1}^{N} u_p\, a_p .}

**Second moment of the numerator.**  With \eqn{c_2 = n(n-1)/[N(N-1)]},

\deqn{\mathbb{E}[Z^2] = \sum_p u_p \frac{n}{N} a_p
  + 2\sum_{p<q} u_p u_q\, c_2\,
    g^{\mathrm{both}}_{pq},}

where, exactly as in Lemma 3,

\deqn{g^{\mathrm{both}}_{pq} =
\begin{cases}
1, & \|x_p-x_q\| \le r,\\
1 - T_{N-2,n-2}(m_p) - T_{N-2,n-2}(m_q)
  + T_{N-2,n-2}(m_p+m_q-w_{pq}), & \text{otherwise}.
\end{cases}}

**Cross moment.**

\deqn{\mathbb{E}[ZY] = \sum_p u_p\frac{n}{N}a_p
  + 2\sum_{p<q}u_pu_q\,c_2\,
    g^{\mathrm{one}}_{pq}, \qquad
  g^{\mathrm{one}}_{pq} =
\begin{cases}
1, & \|x_p-x_q\| \le r,\\
1 - T_{N-2,n-2}(m_p), & \text{otherwise}.
\end{cases}}

The reason for the two cases is that, conditional on both \eqn{p} and \eqn{q}
being marked, a hit for \eqn{p} is already guaranteed when the two are within
\eqn{r}; otherwise at least one of \eqn{p}'s own neighbours must be selected.
Then

\deqn{\mathrm{Cov}(Z,Y) = \mathbb{E}[ZY] - \mathbb{E}[Z]\mathbb{E}[Y].}

**Combining them.**  The package reports the ratio of expectations as the null
mean and a first-order (delta method) variance:

\deqn{\mathbb{E}\!\left[\hat{G}_{\mathrm{rs}}\right] \;\approx\;
  \frac{\mu_Z}{\mu_Y},}

\deqn{\mathrm{Var}\!\left(\hat{G}_{\mathrm{rs}}\right) \;\approx\;
  \frac{V_Z}{\mu_Y^{2}}
  - \frac{2\mu_Z\,\mathrm{Cov}(Z,Y)}{\mu_Y^{3}}
  + \frac{\mu_Z^{2} V_Y}{\mu_Y^{4}},}

where \eqn{\mu_Z = \mathbb{E}[Z]}, \eqn{V_Z = \mathrm{Var}(Z)} and
\eqn{\mu_Y, V_Y} are the denominator moments.  The first two moments of
\eqn{(Z, Y)} are **exact**; only the way they are combined into a mean and
variance for the ratio is approximate, with bias of order \eqn{1/n}.  Radii at
which the population expectation \eqn{\mu_Y} vanishes give no estimate, and are
returned as NA.

## 4. Correction han (Hanisch)

The Hanisch estimator re-weights each event by the area still available at its
radius.  With bin edges \eqn{0 = r_0 < r_1 < \dots} and the eroded-window areas
\eqn{a(r) = |W \ominus r|} (the area of the set of window points with boundary
distance greater than \eqn{r}),

\deqn{\hat{G}_{\mathrm{han}}(r_k) =
  \frac{\sum_{j\le k} h_j / a(r_j)}
       {\max_k \sum_{j\le k} h_j/a(r_j)},}

where \eqn{h_j} counts the uncensored events whose observed distance
\eqn{o_p = \min(\mathrm{nnd}_p, b_p)} falls in bin \eqn{j}.  Equivalently,
\eqn{h_j} counts marked points with \eqn{\mathrm{nnd}_p \in (r_{j-1}, r_j]}
and \eqn{\mathrm{nnd}_p \le b_p}.

Under the null, conditional on \eqn{p \in S}, at least one selected neighbour
within distance \eqn{t} has probability

\deqn{F_p(t) = 1 - T_{N-1,n-1}\big(m_p(t)\big), \qquad
  m_p(t) = |\{q \neq p : \|x_p-x_q\| \le t\}|,}

so the *exact* expected bin counts are

\deqn{\mathbb{E}[h_j] = \frac{n}{N}\sum_{p=1}^{N}
  \Big[ F_p\!\left(\min(r_j, b_p)\right)
      - F_p\!\left(\min(r_{j-1}, b_p)\right) \Big].}

The package computes these expectations exactly and then forms the corrected
curve by plug-in, replacing each \eqn{h_j} with \eqn{\mathbb{E}[h_j]}.  The
expected bin counts are exact; the curve assembled from them is a mean-field
approximation, because \eqn{\mathbb{E}[f(H)] \neq f(\mathbb{E}H)} for the
nonlinear normalisation.  No closed-form variance is available, so csr_var is NA;
use the permutation fallback if a null band for han is needed.

## 5. Correction km (Kaplan-Meier)

The Kaplan-Meier correction treats a point whose nearest neighbour lies beyond
the boundary as right-censored at \eqn{b_p}.  The observed variable is again
\eqn{o_p = \min(\mathrm{nnd}_p, b_p)} with event indicator
\eqn{\delta_p = \mathbf{1}\{\mathrm{nnd}_p \le b_p\}}, and the product-limit
estimate of the survival function is converted to a distribution function:

\deqn{\hat{G}_{\mathrm{km}}(r_k) = 1 - \prod_{j\le k}
  \left(1 - \frac{\mathrm{nco}_j}{d_j}\right),}

where \eqn{\mathrm{nco}_j} is the number of uncensored events in bin \eqn{j}
(so \eqn{\mathbb{E}[\mathrm{nco}_j] = \mathbb{E}[h_j]} above) and
\eqn{d_j} is the number at risk just before bin \eqn{j}.

The expected histogram of \eqn{o_p} is exact as well:

\deqn{\mathbb{E}[\mathrm{obs}_j] = \frac{n}{N}\sum_{p=1}^{N}
  \Big[ F_p(\min(r_j,b_p)) - F_p(\min(r_{j-1},b_p))
      + \mathbf{1}\{r_{j-1} < b_p \le r_j\}\big(1 - F_p(b_p)\big) \Big].}

The last term is the censoring mass: when the boundary distance itself falls in
bin \eqn{j}, then \eqn{o_p = b_p} exactly and no event can occur beyond it.
Setting

\deqn{d_k = \sum_{j \ge k} \mathbb{E}[\mathrm{obs}_j], \qquad
  s_k = 1 - \frac{\mathbb{E}[\mathrm{nco}_k]}{d_k},}

the reported mean-field curve is \eqn{1 - \prod_{j\le k} s_j}.  As with han the
bin expectations are exact and the product-limit assembly is a plug-in, so no
closed-form variance is reported.

## 6. What is exact

| correction | null mean                        | null variance                     | status |
|------------|----------------------------------|-----------------------------------|--------|
| none       | exact (Theorem 1 / Theorem 3)    | exact (Theorem 2 / Theorem 4)     | fully exact |
| rs         | ratio of exact expectations      | delta method from exact moments   | approximate, bias O(1/n) |
| han        | plug-in of exact bin expectations| not available                     | mean-field; permutation for variance |
| km         | plug-in of exact bin expectations| not available                     | mean-field; permutation for variance |

The package records this in the returned object: exact is TRUE only for
correction = "none" with method = "exact".

## 7. Practical guidance

* Because the window built automatically from a data frame is the **convex
  hull** of the points, many locations sit on or near its boundary and
  \eqn{b_p} can be very small.  Edge-corrected analyses are much more
  informative when the user supplies a meaningful window, for example by passing
  a spatstat ppp object with an explicit observation window.
* The reduced-sample estimator is undefined for radii larger than the largest
  \eqn{b_p}; those rows are returned as NA.
* The Hanisch curve is undefined once the eroded window has zero area, and
  those rows are NA as well.
* If a null band is required for han or km, use permute_gest with the same
  correction.

## 8. Numerical check: rs against exhaustive enumeration

With an explicit unit-square window, every marked subset can be enumerated for a
small pattern and compared with the closed-form reduced-sample moments.

```{r rs-check, message = FALSE}
suppressPackageStartupMessages(library(permfreeG))
set.seed(3)
N <- 10; n <- 4
d <- data.frame(x = stats::runif(N), y = stats::runif(N))
d$m <- factor(c(rep("A", n), rep("B", N - n)))
pp <- spatstat.geom::ppp(d$x, d$y,
                         window = spatstat.geom::owin(c(0, 1), c(0, 1)),
                         marks = d$m)
radii <- seq(0, 0.3, by = 0.05)

D <- as.matrix(stats::dist(cbind(d$x, d$y)))
bd <- spatstat.geom::bdist.points(pp)
ci <- utils::combn(N, n)
est <- matrix(NA_real_, ncol(ci), length(radii))
for (j in seq_len(ncol(ci))) {
  S <- ci[, j]
  nnd <- apply(D[S, S], 1, function(z) min(z[z > 0]))
  for (k in seq_along(radii)) {
    den <- sum(bd[S] > radii[k])
    est[j, k] <- if (den > 0) {
      sum(nnd <= radii[k] & bd[S] > radii[k]) / den
    } else NA_real_
  }
}
ok <- colSums(!is.na(est)) == ncol(ci)
res <- as.data.frame(exact_gest(pp, radii = radii, marks_i = "A",
                                correction = "rs"))
data.frame(r = radii[ok],
           enumerated_mean = colMeans(est[, ok, drop = FALSE]),
           exact_rs_mean   = res$csr_mean[ok],
           enumerated_var  = apply(est[, ok, drop = FALSE], 2,
                                   function(z) mean((z - mean(z))^2)),
           delta_var       = res$csr_var[ok])
```

The ratio of expectations reproduces the enumerated mean to within a fraction
of a percent, and the delta-method variance is close to but not identical with
the enumerated variance, as expected.

## 9. Numerical check: han and km against permutation

On a window with mild censoring, the mean-field curves are close to the
permutation average.

```{r han-km-check, message = FALSE}
set.seed(99)
N <- 40; n <- 15
d2 <- data.frame(x = stats::runif(N), y = stats::runif(N))
d2$m <- factor(c(rep("A", n), rep("B", N - n)))
pp2 <- spatstat.geom::ppp(d2$x, d2$y,
                          window = spatstat.geom::owin(c(0, 1), c(0, 1)),
                          marks = d2$m)
radii <- seq(0, 0.4, by = 0.05)

out <- lapply(c("han", "km"), function(cor) {
  ex <- as.data.frame(exact_gest(pp2, radii = radii, marks_i = "A",
                                 correction = cor))
  pm <- permute_gest(pp2, radii = radii, marks_i = "A", correction = cor,
                     nsim = 200)
  data.frame(correction = cor, r = radii,
             mean_field = ex$csr_mean, permutation_mean = pm$mean)
})
out <- do.call(rbind, out)
aggregate(abs(out$mean_field - out$permutation_mean),
          by = list(correction = out$correction), FUN = max)
```

The mean-field approximation is adequate here (differences of a few percent) but
should not be trusted when the marked count is very small or when most points are
heavily censored; in those regimes the permutation fallback is the right tool.
