---
title: "Getting started with permfreeG"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with permfreeG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4)
```

## What the package does

permfreeG computes, in closed form, the mean and variance of the
nearest-neighbour \eqn{G} function under the mark-permutation null.  Instead of
simulating thousands of permutations, it evaluates the exact moments directly,
so a null band and pointwise z-scores are available immediately.

Two statistics are covered:

* exact_gest, the univariate nearest-neighbour \eqn{G};
* exact_gcross, the bivariate cross nearest-neighbour \eqn{G_{ij}}.

Both accept a data frame with columns x, y (and optionally a mark column) or an
existing spatstat ppp object.

## Installation

```{r install, eval = FALSE}
# development version (the package lives on the dev branch)
remotes::install_github("ACSoupir/PermFree-NearestNeighborG", ref = "dev",
                        build_vignettes = TRUE)
```

On macOS, if your R toolchain picks up a non-Apple compiler from the PATH,
prepend /usr/bin before installing, or place an appropriate ~/.R/Makevars in
your home directory.

## A first univariate example

The package ships sim_nnG, a 30-point pattern with five marked points.  This is
the small example in which all choose(30, 5) = 142506 marked subsets could be
enumerated by hand; the closed form must agree with their average.

```{r univariate, message = FALSE}
suppressPackageStartupMessages(library(permfreeG))
str(sim_nnG)

radii <- seq(0, 1.2, by = 0.15)
res <- exact_gest(sim_nnG, radii = radii, mark_col = "m", marks_i = "A")
res
res_df <- as.data.frame(res)
```

The returned object has one row per radius:

* r, the radius;
* theo, the theoretical CSR curve \eqn{G(r) = 1 - e^{-\lambda\pi r^2}} using the
  observed intensity;
* obs, the observed curve for the marked points;
* csr_mean and csr_var, the exact null mean and variance from Theorems 1-2;
* z and p_value, pointwise departures of the observed curve from the exact null.

```{r univariate-plot, fig.height = 4, fig.alt = "Observed G curve with the exact CSR mean and a one standard deviation band."}
plot(res)
```

Because csr_var is exact rather than estimated, z has its usual interpretation
at every radius simultaneously; no permutation noise inflates it.

## Bivariate cross G

```{r bivariate, message = FALSE}
set.seed(7)
d3 <- data.frame(x = stats::runif(60), y = stats::runif(60),
                 m = sample(c("A", "B", "C"), 60, replace = TRUE))
gc <- exact_gcross(d3, radii = seq(0, 1.5, by = 0.15), mark_col = "m",
                   marks_i = "A", marks_j = "B")
gc
```

Passing the same mark twice (marks_i == marks_j) reduces to exact_gest with
n equal to the observed count of that mark.

## Edge corrections

Corrections are selected with correction.  The raw curve is exact; the border
(reduced sample) option uses a delta method built from exact moments, and han/km
are mean-field curves with no closed-form variance.

```{r corrections}
pp <- spatstat.geom::ppp(d3$x, d3$y,
                         window = spatstat.geom::owin(c(0, 1), c(0, 1)),
                         marks = d3$m)
rr <- seq(0, 0.4, by = 0.05)

none_cur <- as.data.frame(exact_gest(pp, radii = rr, marks_i = "A"))
rs_cur   <- as.data.frame(exact_gest(pp, radii = rr, marks_i = "A",
                                     correction = "rs"))
data.frame(r        = none_cur$r,
           raw_mean = round(none_cur$csr_mean, 4),
           rs_mean  = round(rs_cur$csr_mean, 4),
           raw_var  = signif(none_cur$csr_var, 3),
           rs_var   = signif(rs_cur$csr_var, 3))
```

The border correction is undefined beyond the largest boundary distance; those
rows are NA.  See the "Supplement: edge corrections" vignette for the full
derivation of each option, including which parts are exact.

## Approximations when speed matters

Two knobs trade accuracy for time:

* covariance = FALSE drops the off-diagonal covariance terms, keeping only
  \eqn{\sum_p e_p(1-e_p)/n^2}.  This is the independent-points approximation and
  costs \eqn{O(RN)} once neighbour counts are known.
* method = "approx" is the same thing expressed through the public interface.

```{r approx}
fast <- exact_gest(sim_nnG, radii = radii, mark_col = "m", marks_i = "A",
                   method = "approx")
c(exact_variance    = sum(res_df$csr_var),
  approx_variance   = sum(as.data.frame(fast)$csr_var))
```

## Permutation fallback

Nothing is hidden: the exact moments can always be checked against a real
permutation run.  permute_gest and permute_gcross draw uniform subsets (or
disjoint label permutations) and recompute the statistic, optionally in
parallel.

```{r permutation, message = FALSE}
set.seed(1)
pm <- permute_gest(sim_nnG, radii = radii, mark_col = "m", marks_i = "A",
                   nsim = 200)
data.frame(r          = radii,
           exact_mean = round(res_df$csr_mean, 4),
           perm_mean  = round(pm$mean, 4),
           exact_var  = signif(res_df$csr_var, 3),
           perm_var   = signif(pm$var, 3))
```

The two agree up to Monte Carlo error.  The permutation route is the reference
implementation, and it also supplies a variance for the han/km corrections,
where no closed form is available.

## Scaling

The exact engine stores neighbour sets as packed bit sets, so each pair
intersection costs one word of population count rather than a set intersection.
The result is that the whole variance curve for N = 250 at 151 radii takes a
fraction of a second.

```{r scaling}
set.seed(1)
N <- 250
x <- stats::runif(N); y <- stats::runif(N)
grid <- seq(0, 1.5, length.out = 151)

t_exact <- system.time(
  permfreeG:::csr_uni_moments_cpp(x, y, grid, 50L))[["elapsed"]]
t_approx <- system.time(
  permfreeG:::csr_uni_moments_cpp(x, y, grid, 50L,
                                  covariance = FALSE))[["elapsed"]]
c(N = N, radii = length(grid), exact_seconds = t_exact,
  approx_seconds = t_approx)
```

## Where the theory lives

* "Derivation: univariate Gest" proves Theorems 1-2 with full proofs.
* "Derivation: bivariate Gcross" proves the conditioning lemma and Theorems 3-4.
* "Supplement: edge corrections" derives none, rs, han and km.

No code is required to follow any of the three.
